\documentclass[xcolor={dvipsnames}]{beamer}

\usetheme{metropolis}

\usepackage{ebproof}
\usepackage{wasysym}
\usepackage{tikz}
\usepackage{tikz-cd}
\usetikzlibrary{tikzmark}
\usetikzlibrary{calc}
\usepackage[backend=biber]{biblatex}
\addbibresource{references.bib}

\title{Formal Topology in Univalent Foundations}
\date{\today}
\author{Ayberk Tosun}
\institute{Chalmers University of Technology}

\setmonofont{PragmataPro Mono Liga}

\definecolor{AgdaFunction}{HTML}{0000CD}
\definecolor{AgdaString}{HTML}{B22222}
\definecolor{codecolour}{RGB}{27, 134, 236}
\definecolor{gerbesred}{RGB}{164, 30, 50}
\definecolor{darkred}{rgb}{0.45,0,0}

\definecolor{darkgreen}{rgb}{0,0.45,0}
\definecolor{hottviolet}{rgb}{0.45,0,0.45}
\definecolor{hottblue}{rgb}{0,0.45,0.45}

\hypersetup{
  linktoc    = page,
  citecolor  = hottblue,
  urlcolor   = hottviolet
}

\newcommand{\fnname}[1]{{\color{codecolour} {\tt #1}}}
\newcommand{\prgoutput}[1]{{\color{codecolour} {\tt #1}}}

\newcommand{\pity}[3]{\prod_{(#1~:~#2)} #3}
\newcommand{\sigmaty}[3]{\sum_{(#1~:~#2)} #3}
\newcommand{\pow}[1]{\mathcal{P}\left(#1\right)}
\newcommand{\univ}{\mathsf{Type}}
\newcommand{\trunc}[1]{\left\| #1 \right\|}
\newcommand{\abs}[1]{\left| #1 \right|}
\newcommand{\nuclnm}{\mathbf{j}}
\newcommand{\nuclapp}[1]{\nuclnm{}\left(#1\right)}

\newcommand{\fix}[2]{\mathfrak{fix}\left(#1, #2\right)}

\newcommand{\rulename}[1]{{\color{darkred} \mathsf{#1}}}
\newcommand{\ruledir}{{\color{gerbesred} \mathsf{dir}}}
\newcommand{\rulebranch}{{\color{gerbesred} \mathsf{branch}}}
\newcommand{\rulesquash}{{\color{gerbesred} \mathsf{squash}}}
\newcommand{\hprop}{\mathsf{hProp}}

\newcommand{\McF}{\mathcal{F}}

\newcommand{\is}{:\equiv}

\newcommand{\covers}[2]{#1 \lhd #2}
\newcommand{\badcovers}[2]{#1~\LHD~#2}

%% Color customisation.

\begin{document}

\maketitle

\section{What is topology?}

\begin{frame}{What is topology?}
  \large
  A topological space is a set $X$ together with a collection $\Omega(X)$ of its subsets
  such that
  \begin{itemize}
    \item $\Omega(X)$ is closed under \alert{finite} intersections, and
    \item $\Omega(X)$ is closed under \alert{arbitrary} unions.
  \end{itemize}
\end{frame}

\begin{frame}[c]{What is topology?}
  \large
  \vspace{2em}

  Let \fnname{P} be a program. When run, it prints a sequence of integers.

  \vspace{1em}

  We \alert{observe} that its output starts as:

  \begin{center}
  \prgoutput{7~~11~~2~~2~~8~~42~~}~$\cdots$
  \end{center}

  We can consider certain properties of \fnname{P}, such as:
  \begin{center}
    ``\tikzmarknode{a}{\fnname{P}} eventually prints \fnname{17}'', or\\
    \vspace{1em}
    ``\fnname{P} prints no more than two \fnname{2}s\tikzmarknode{b}{''}.
  \end{center}

  \uncover<2>{
    {\color{gerbesred}
      \vspace{1em}
      \tikzmarknode{txt}{Observable}%
      \hspace{14.4em}%
      \tikzmarknode{not-obs}{\textbf{Not} observable}
    \begin{tikzpicture}[remember picture, overlay]
      \draw[->, line width=0.3mm, shorten >=3mm] (txt.north) to[out=90, in=-180] (a.west);
      \draw[->, line width=0.3mm, shorten >=3mm] (not-obs.north) to[out=80, in=0] (b.east);
    \end{tikzpicture}
    }
  }
\end{frame}

\begin{frame}{What is topology?}
  \large
  \begin{center}
    ``$\phi$ is an \alert{observable property}.''

    \vspace{0.5em}
    $\leftrightarrow$
    \vspace{0.5em}

    If a program satisfies $\phi$, there exists a stage $i$ of the output $\sigma$ at which the
    program is \alert{verified} to satisfy $\phi$: all extensions of $\sigma|_i$ satisfy $\phi$.
  \end{center}
\end{frame}

\begin{frame}{What is topology?}
  Let $\phi_1, \cdots ,\phi_n$ be a \alert{finite} number of observable properties.

  \vspace{1em}

  Suppose $\phi_1 \wedge \cdots \wedge \phi_n$ holds.

  \vspace{1em}

  There must be \alert{stages} $i_1, \cdots , i_n$ such that $\phi_k$ is verified at stage $i_k$.

  \vspace{1em}

  $\phi_1 \wedge \cdots \wedge \phi_n$ must then be verified at $\max(i_1, \cdots, i_n)$.

  \vspace{1em}

  This means that $\phi_1 \wedge \cdots \wedge \phi_n$ is \alert{observable}.
\end{frame}

\begin{frame}{What is topology?}
    Let $\{~\psi_i ~|~ i \in I~\}$ be an \alert{arbitrary} number of observable properties.

    \vspace{1em}

    Suppose $\bigvee_i \psi_i$ holds.

    \vspace{1em}

    Some $\psi_i$ holds meaning it must be verified at some stage $m$.

    \vspace{1em}

    $\bigvee_i \psi_i$ is hence verified at stage $m$.

    \vspace{1em}

    This means that $\bigvee_i \psi_i$ is \alert{observable}.
\end{frame}

\begin{frame}{What is topology?}
  \huge
  \begin{center}
    Topology is a mathematical theory of \alert{observable} properties.
  \end{center}
\end{frame}

\section{Frames}

\begin{frame}{Frames}
  \large
  \vspace{1.5em}

  A \alert{frame} is a poset $\mathcal{O}$ such that
  \begin{itemize}
    \item \alert{finite subsets} of $\mathcal{O}$ have \alert{meets},
    \item \tikzmarknode{a}{\alert{arbitrary subsets}} of $\mathcal{O}$ have \alert{joins},
      and
    \item binary meets distribute over arbitrary joins:
      \begin{equation*}
        a \wedge \left( \bigvee_{i~\in~I} b_i \right) = \bigvee_{i~\in~I} \left( a \wedge b_i \right),
      \end{equation*}
      for any $a \in \mathcal{O}$ and family $\{ b_i ~|~ i \in I \}$ over $\mathcal{O}$.
  \end{itemize}

  \vspace{1em}

  \uncover<2>{
    {\color{gerbesred}
    \begin{center}
    {\Large
      \tikzmarknode{b}{In} type theory, the quantification over arbitrary subsets is
      problematic.
    }
    \end{center}
    \begin{tikzpicture}[remember picture, overlay]
      \draw[->, line width=0.3mm, shorten >=5mm] (b.west) to[out=180, in=-180] (a.west);
    \end{tikzpicture}
    }
  }
\end{frame}

\begin{frame}{Frames --- a prime example}
  \large

  Given a poset
  \begin{align*}
    A &\quad:\quad \univ{}_m\\
    \sqsubseteq &\quad:\quad A \rightarrow A \rightarrow \mathsf{hProp}_m
  \end{align*}
  the type of \alert{downwards-closed subsets} of $A$ is:
  \[ \sigmaty{U}{\pow{A}}{\pity{x~y}{A}{x \in U \rightarrow y \sqsubseteq x \rightarrow y \in U}}, \]
  \begin{center}
    where
  \end{center}
  \begin{alignat*}{2}
    \mathcal{P}    \quad&:\quad   && \univ{}_m \rightarrow \univ{}_{m+1} \\
    \mathcal{P}(X) \quad&\is\quad && X \rightarrow \mathsf{hProp}_m     . 
  \end{alignat*}
\end{frame}

\begin{frame}{Frames --- a prime example}
  \large

  This forms a \alert{frame} defined as:
  \begin{align*}
    \top           &\quad\is{}\quad \lambda \_.~ \mathsf{Unit}\\
    A \wedge B       &\quad\is{}\quad \lambda x.~ (x \in A) \times (x \in B)\\
    \bigvee_{i~:~I} B_i &\quad\is{}\quad \lambda x.~ \trunc{\sigmaty{i}{I}{x \in B_i}}.
  \end{align*}
\end{frame}

\begin{frame}{Frames --- a prime example}
  \large

  \textbf{Question}: can we get any frame out of a poset in this way?

  \vspace{1em}

  \uncover<2>{
    One way is to employ the notion of a \alert{nucleus} on a frame.

    \vspace{1em}

    For this, we need to enrich the notion of a poset with a structure that gives rise to
    an appropriate \alert{nucleus}\\
    (on its frame of downwards-closed subsets).

    \vspace{1em}

    That structure is a \alert{formal topology}.
  }
\end{frame}

\section{Formal Topology}

\begin{frame}{Formal Topologies --- as Interaction Systems}
  \large

  An \alert{interaction structure}~\cite{tree-sets} on some type $A$ comprises three
  functions:
  \begin{align*}
    B  &\quad:\quad A \rightarrow \univ{}                               && \text{(1)},           \\
    C  &\quad:\quad \pity{a}{A}{B(a) \rightarrow \univ{}}               && \text{(2)}, \text{and}\\
    d  &\quad:\quad \pity{a}{A}{\pity{b}{B(a)}{C(a, b) \rightarrow A}}  && \text{(3)}.
  \end{align*}

  An \alert{interaction system} is a type $A$ equipped with an interaction structure.
\end{frame}

\begin{frame}{Formal Topologies --- as Interaction Systems}
  \large

  A \alert{formal topology} is an interaction system $(B, C, d)$ on some poset $P$ that
  satisfies the following two conditions.

  \begin{enumerate}
    \item \textbf{Monotonicity}:
      {\normalsize
      \begin{equation*}
        \pity{a}{A}{\pity{b}{B(a)}{\pity{c}{C(a, b)}{d(a, b, c) \sqsubseteq a}}}.
      \end{equation*}
      }
    \item \textbf{Simulation}:
      {\normalsize
      \begin{align*}
        &\pity{a'~a}{A}{
          a' \sqsubseteq a \rightarrow\\
            &\quad\pity{b}{B(a)}{%
              \sigmaty{b'}{B(a')}{
                \pity{c'}{C(a', b')}{
                  \sigmaty{c}{C(a, b)}{d(a', b', c') \sqsubseteq d(a, b, c)}
                }
              }
            }
        }.
      \end{align*}
      }
  \end{enumerate}
\end{frame}

\begin{frame}{Nuclei}
  \large
  A \alert{nucleus} on a frame $F$ is an endofunction $\nuclnm : \abs{F} \rightarrow \abs{F}$ such
  that:
  \begin{alignat*}{4}
    &\pity{x~y}{\abs{F}}{~&& \nuclapp{x \wedge y}       \quad&&=\quad \nuclapp{x} \wedge \nuclapp{y}}
      &&\qquad\text{[meet preservation],}\\
    &\pity{x~~}{\abs{F}}{~&& x                     \quad&&\sqsubseteq\quad \nuclapp{x}}
      &&\qquad\text{[inflation], and}\\
    &\pity{x~~}{\abs{F}}{~&& \nuclapp{\nuclapp{x}} \quad&&\sqsubseteq\quad \nuclapp{x}}
      &&\qquad\text{[idempotence].}
  \end{alignat*}

  \vspace{0.5em}

  \uncover<2>{\Large
    \begin{center}
      This is a meet-preserving, \alert{idempotent monad}!
    \end{center}
  }
\end{frame}

\begin{frame}{Nuclei}
  \large

  Let $F$ be a frame, and $\nuclnm{} : \abs{F} \rightarrow \abs{F}$ a nucleus on it.

  \vspace{1em}

  The set
  \begin{equation*}
    \{ x \in \abs{F} ~|~ \nuclapp{x} = x \}
  \end{equation*}
  of \alert{fixed points} for $\nuclnm{}$ is itself a frame:
  \begin{align*}
    \top       \quad&\is\quad \top_F                \\
    \_\wedge\_   \quad&\is\quad \_\wedge_F\_            \\
    \bigvee_i x_i \quad&\is\quad \nuclapp{\bigvee^F_i x_i} .
  \end{align*}

  We denote this $\fix{F}{\nuclnm{}}$.
\end{frame}

\begin{frame}{Nuclei}
  \large

  \hrule
  \begin{quote}
    A Grothendieck ``topology'' appears most naturally as a modal operator, of the nature
    ``it is locally the case that''.
    \begin{flushright}
      \emph{--- Lawvere~\cite{quantifiers-and-sheaves}}
    \end{flushright}
  \end{quote}
  \hrule

  \vspace{1.5em}

  \uncover<2>{
    In the posetal case, our \emph{modality} will be the \alert{covering} relation induced
    by the structure of a formal topology.
  }
\end{frame}

\begin{frame}{The covering nucleus --- naive attempt}
  \large

  Let\\
  \begin{itemize}
    \item $\McF{}$ be a formal topology with underlying poset $P$,\\
    \item $a : \abs{P}$, and\\
    \item $U : \pow{\abs{P}}$, a downwards-closed subset of $P$.
  \end{itemize}

  \vspace{\baselineskip}

  $\covers{a}{U}$ is inductively defined via two rules.

  \[
    \begin{prooftree}
      \hypo{ a \in U }
      \infer1[$\ruledir{}$]{\covers{a}{U}}
    \end{prooftree}
    \qquad
    \begin{prooftree}
      \hypo{b : B(a)}
      \hypo{\pity{c}{C(a, b)}{\covers{d(a, b, c)}{U}}}
      \infer2[$\rulebranch{}$]{\covers{a}{U}}
    \end{prooftree}
  \]

  \vspace{1em}

  \uncover<2>{%
    \emph{Notice: $\covers{a}{U}$ is a \textbf{structure} and not a \textbf{property}.}
  }
\end{frame}

\begin{frame}{The covering nucleus --- naive attempt}
  \large

  $\rhd$ could be shown to be a \alert{nucleus}, \textbf{if it had the type}
  \begin{align*}
    \lhd \quad&:\quad \abs{P} \rightarrow \pow{\abs{P}} \rightarrow \hprop{} \\
    \rhd \quad&:\quad \pow{\abs{P}} \rightarrow \pow{\abs{P}},
  \end{align*}
  but its type is
  \begin{align*}
    \lhd \quad&:\quad \abs{P} \rightarrow \pow{\abs{P}} \rightarrow \univ{}.
  \end{align*}

  \uncover<2>{%
    \textbf{Idea}: use propositional truncation:
    \begin{align*}
      \trunc{\_ \lhd \_} &: \abs{P} \rightarrow \pow{\abs{P}} \rightarrow \hprop{} \\
      \trunc{\_ \rhd \_} &: \pow{\abs{P}} \rightarrow \pow{\abs{P}}      .
    \end{align*}
  }
\end{frame}

\begin{frame}{The covering nucleus --- naive attempt}
  \large

  \textbf{Need to show}: $\trunc{\_ \lhd \_}$ is a nucleus.

  \vspace{1em}

  This involves showing it is idempotent:
  \begin{equation*}
    \trunc{\_ \lhd \trunc{\_ \lhd U}} \quad\subseteq\quad \trunc{\_ \lhd U},
  \end{equation*}
  for which we need to prove a lemma stating:
  \begin{equation*}
    \covers{a}{U}
    \times
    \left(\pity{u}{\abs{P}}{u \in U \rightarrow \trunc{\covers{u}{V}}\right) \rightarrow \trunc{\covers{a}{V}}},
  \end{equation*}
  \begin{quote}
    {\scriptsize
    for every formal topology $\McF{}$ with underlying poset $P$, $a : \abs{P}$, and
    downwards-closed subsets $U, V : \pow{\abs{P}}$.
    }
  \end{quote}
\end{frame}

\begin{frame}{The covering nucleus --- naive attempt}
  In the $\rulebranch{}$ case of an attempted proof, the inductive hypothesis gives us
  \begin{equation*}
    \pity{c}{C(a, b)}{\trunc{\covers{d(a, b, c)}{V}}},
  \end{equation*}
  but what we need is:
  \begin{equation*}
    \trunc{\pity{c}{C(a, b)}{\covers{d(a, b, c)}{V}}}.
  \end{equation*}

  \vspace{1em}

  \uncover<2->{
    \begin{center}
      \Large
      This would require (a form of) the axiom of choice.

      In fact, the form of choice needed is provably false~\cite[Lemma 3.8.5]{hottbook}.
    \end{center}
  }
\end{frame}

\begin{frame}{The covering nucleus --- fixed}
  As we cannot truncate, we \emph{revise} the inductive definition of $\lhd$ to be a
  \alert{higher inductive type}.

  \vspace{\baselineskip}

  \hrule
  \begin{equation*}
    {
      \begin{prooftree}
        \hypo{ a \in U }
        \infer1[$\ruledir{}$]{\covers{a}{U}}
      \end{prooftree}
    }
    \qquad
    {
      \begin{prooftree}
        \hypo{b : B(a)}
        \hypo{\pity{c}{C(a, b)}{\covers{d(a, b, c)}{U}}}
        \infer2[$\rulebranch{}$]{\covers{a}{U}}
      \end{prooftree}
    }
  \end{equation*}
  \vspace{1em}
  \begin{equation*}
    {
      \begin{prooftree}
        \hypo{p : \covers{a}{U}}
        \hypo{q : \covers{a}{U}}
        \infer2[$\rulesquash{}$]{p = q}
      \end{prooftree}
    }
  \end{equation*}
  \hrule

  \uncover<2>{%
    \begin{center}
      \Large
      The mentioned lemma is now provable without choice \emph{and} the type is
      propositional!
    \end{center}
  }
\end{frame}

\begin{frame}{Generating frames from formal topologies}
  \large

  \begin{enumerate}
    \item Start with formal topology $\McF{}$ with underlying poset $P$.
    \item Take the frame of downwards-closed subsets of $P$, denoted $P\downarrow$.
    \item $\rhd : P\downarrow \rightarrow P\downarrow$ is a nucleus.
    \item The generated frame is the \alert{frame of fixed points} of this nucleus
      (denoted $\mathfrak{fix}\left(P\downarrow, \rhd\right)$).
  \end{enumerate}
\end{frame}

\section{Formal topologies present}

\begin{frame}{Flat monotonic maps}
  To state the presentation theorem, we will have to talk about meet-preserving monotonic
  maps between posets.

  \vspace{0.5em}

  However, we are working with posets which may or may not have meets.

  \vspace{0.5em}

  The solution is to consider those monotonic maps preserving \alert{latent meets}: these
  are called \alert{flat monotonic maps}.

  \vspace{0.5em}

  Let $f : P \rightarrow F$ be a \textbf{monotonic map} from a poset $P$ to the underlying poset of
  a frame $F$. We say that it is \alert{flat} if:
  \begin{alignat*}{4}
    & &&\top_F &&\quad=\quad &&\bigvee \{ f(a) ~|~ a : \abs{P} \},\quad\text{and}\\
    &\pity{a_0~a_1}{\abs{P}}{&&f(a_0) \wedge f(a_1) &&\quad=\quad &&\bigvee \{ f(a_2) ~|~ a_2 \sqsubseteq a_0 ~\text{and}~ a_2 \sqsubseteq a_1 \} }.
  \end{alignat*}
\end{frame}

\begin{frame}{Representation}
  Let
  \begin{itemize}
    \item $\McF{}$ be a formal topology,
    \item $R$, a frame, and
    \item $f : \abs{\McF{}} \rightarrow \abs{R}$, a function.
  \end{itemize}
  

  \vspace{1em}

  We say that $f$ \alert{represents} $\McF{}$ in $R$ if:
  \begin{equation*}
    \pity{a}{A}{\pity{b}{B(a)}{f(a) \sqsubseteq \bigvee_{c : C(a, b)} f(d(a, b, c))}}.
  \end{equation*}
\end{frame}

\begin{frame}[fragile]{The main theorem}
  \textbf{Theorem.} Given
  \begin{itemize}
    \item a formal topology $\McF{}$ with underlying poset $P$,
    \item a frame $R$, and
    \item a \textbf{flat} monotonic map $f : P \rightarrow R$;
  \end{itemize}
  \begin{center}
    if $f$ represents $\McF{}$ in $R$, then there exists a \textbf{unique} frame
    homomorphism $g$ making the following diagram commute:
  \end{center}

  \begin{center}
    \begin{tikzcd}[row sep=40pt, column sep=40pt]\label{fig:comm-diag}
      P \arrow[swap, rd, "f"] \arrow[r, "\eta"]
        & \fix{P\downarrow}{\rhd} \arrow[d, dashed, "g"] \\
        & R
    \end{tikzcd}\\
    \vspace{0.5em}
    {\footnotesize where $\eta(a) \is \covers{\_}{ \{ a' ~|~ a' \sqsubseteq a \} }$.}
  \end{center}
\end{frame}

\begin{frame}{Closing remarks}
  Conclusion.
  \begin{itemize}
    \item We have reconstructed the notion of a covering within the univalent doctrine as
      an HIT.
    \item We have sketched the beginnings of an approach for carrying out formal topology
      in univalent type theory.
    \item No postulates, no setoids, no impredicativity.
  \end{itemize}
\end{frame}

\begin{frame}{Further work}
  Further work.
  \begin{itemize}
    \item Develop more topology this way!
    \item What is the category of formal topologies?
    \item How can the presentation theorem be stated as an adjunction?
  \end{itemize}
\end{frame}


\section{QEF}

\begin{frame}[noframenumbering]{References}
  \printbibliography[heading=none]
\end{frame}

\end{document}
